---
title: "Statistical Inference Project - Part I"
subtitle: "Exploration of the exponential distribution"
author: "Loïc BERTHOU"
date: "August 18, 2015"
output: pdf_document
---

```{r, echo=FALSE, results='hide'}
set.seed(20150815)
library(ggplot2)
```

# Overview

This document looks at the behaviour of the exponential distribution, specifically its mean distribution and variance distribution for a set of samples. The observations will help understand the Central Limit Theorem.

# Introduction

For this study, we will investigate the distribution of averages of 40 exponentials that will be simulated 1000 times. The rate parameter $\lambda$ is set to $\frac{1}{5}$ for all simulations.

# Simulations

First, we generate the 1000 samples of 40 exponentials and store them in the variable _expMatrix_.

```{r}
n <- 40
lambda <- 1/5
nbSim <- 1000
expMatrix <- matrix(rexp(n*nbSim, lambda), nbSim)
```

# Sample Mean versus Theoretical Mean

The theoretical mean of the exponential distribution (our population) is $\mu = \frac{1}{\lambda}$. In our case, this means that $\mu$ = `r 1/lambda`.

We calculate the mean of each sample to obtain a distribution of 1000 means.

```{r}
expMeans <- apply(expMatrix, 1, mean)
```

We can visualize the distribution of these means on a histogram in *Figure 1*. The theoretical mean has been represented on this figure by the vertical red line.

```{r, fig.width=7, fig.height=4, echo=FALSE}
expMeansFrm <- data.frame(meanSim=expMeans)

g <- ggplot(expMeansFrm, aes(x=meanSim))
g <- g + geom_histogram(aes(y=..density..), alpha = .30, binwidth=0.2, linetype="blank", fill=hcl(220, 50, 70))
g <- g + geom_density(size = 1, color = hcl(220, 50, 70))
g <- g + geom_vline(xintercept = 1/lambda, size = 1, color = hcl(0, 50, 70))
g <- g + geom_vline(xintercept = 0, size = .5, color = "black")
g <- g + geom_hline(yintercept = 0, size = .5, color = "black")
g <- g + ggtitle("Figure 1 - Sample mean distribution of an exponential distribution")
g <- g + labs(x="mean value")
g
```

We observe that the sample mean is distributed around the theoretical mean of the population $\mu$. Let's calculate the mean of this sample distribution and compare it to $\mu$.

```{r, results='hide'}
expMeansMean <- mean(expMeans)
```

The value _expMeansMean_ = `r expMeansMean` is very close to the theoretical mean $\mu$ = `r 1/lambda`. We can then conclude that this demonstrates that the sample mean is approximately the popultation mean.

# Sample Variance versus Theoretical Variance

The theoretical variance of the exponential distribution (our population) is $\sigma^2 = \frac{1}{\lambda^2}$. This means that the theoretical variance of the sample distribution of averages (or Standard Error) is $\frac{\sigma^2}{n}$. Let's calculate the variance of this sample distribution and compare it to the theoretical variance.

```{r, results='hide'}
expMeansVar <- var(expMeans)
```

The value _expMeansVar_ = `r expMeansVar` is very close to the theoretical variance $\frac{\sigma^2}{n}$ = `r (1/lambda^2)/n`. We can then conclude that this demonstrates that the sample variance is approximately the Standard Error of the mean.

# Distribution

As we could see in the figure 1, the distribution of means looks very much like a normal distribution. We will normalize the values obtained previously to be able to compare them with a standard normal curve. To do so we will use the formula:
$\frac{estimate - mean of estimate}{Std Err of estimate}$

```{r}
cfunc <- function(x) (mean(x) - 1/lambda) * sqrt(n) * lambda
expMeansNorm = apply(expMatrix, 1, cfunc)
```

We can visualize the distribution of these normalized means on a histogram in *Figure 3*.
The standard normal distribution has been represented on this figure by the red curve.

```{r, fig.width=7, fig.height=4, echo=FALSE}
expMatrixNormFrm <- data.frame(meanSimNorm=expMeansNorm)

g <- ggplot(expMatrixNormFrm, aes(x=meanSimNorm))
g <- g + geom_histogram(aes(y=..density..), alpha = .30, binwidth=0.2, linetype="blank", fill=hcl(220, 50, 70))
g <- g + geom_density(size = 1, color = hcl(220, 50, 70))
g <- g + stat_function(fun = dnorm, size = 1, color = hcl(0, 50, 70))
g <- g + geom_vline(xintercept = 0, size = .5, color = "black")
g <- g + geom_hline(yintercept = 0, size = .5, color = "black")
g <- g + ggtitle("Figure 2 - Normalized sample mean distribution")
g <- g + labs(x="normalized mean value")
g
```

We observe that the sample mean distribution is extremely close to the standard normal function.
We can then conclude that the distribution is approximately normal.

# Conclusion

We have demonstrated that the Central Limit Theorem does apply to the exponential distribution. 
The distribution of averages for an exponential distribution is approximately a normal distribution with:

- mean equals to the mean of the population.
- variance equals to the standard error of the mean.

$\pagebreak$

# Appendix

## Exponential Distribution

```{r, fig.width=7, fig.height=4}
expValues <- expMatrix
dim(expValues) <- NULL

expValuesFrm <- data.frame(expValues=expValues)

g <- ggplot(data.frame(x = c(0, 40)), aes(x))
g <- g + stat_function(fun = dexp, args = list(rate = lambda), size = .5, color = hcl(180, 50, 70))
g <- g + geom_vline(xintercept = 0, size = .5, color = "black")
g <- g + geom_hline(yintercept = 0, size = .5, color = "black")
g <- g + ggtitle("Figure 3 - exponential distribution")
g <- g + labs(x=NULL)
g

```

$\pagebreak$

## Code for generating Figure 1

```{r, eval=FALSE}
expMeansFrm <- data.frame(meanSim=expMeans)

g <- ggplot(expMeansFrm, aes(x=meanSim))
g <- g + geom_histogram(aes(y=..density..), alpha = .30, binwidth=0.2, linetype="blank")
g <- g + geom_density(size = .5, alpha = .2, colour="blue")
g <- g + geom_vline(xintercept = 1/lambda, size = .5, color = "red")
g <- g + geom_vline(xintercept = 0, size = .5, color = "black")
g <- g + geom_hline(yintercept = 0, size = .5, color = "black")
g <- g + ggtitle("Figure 1 - Sample mean distribution of an exponential distribution")
g <- g + labs(x="mean value")
g
```

## Code for generating Figure 2

```{r, eval=FALSE}
expMatrixNormFrm <- data.frame(meanSimNorm=expMeansNorm)

g <- ggplot(expMatrixNormFrm, aes(x=meanSimNorm))
g <- g + geom_histogram(aes(y=..density..), alpha = .30, binwidth=0.2, linetype="blank")
g <- g + geom_density(size = .5, alpha = .2, color = "blue")
g <- g + stat_function(fun = dnorm, size = .5, color = "red")
g <- g + geom_vline(xintercept = 0, size = .5, color = "black")
g <- g + geom_hline(yintercept = 0, size = .5, color = "black")
g <- g + ggtitle("Figure 2 - Normalized sample mean distribution")
g <- g + labs(x="normalized mean value")
g
```
