---
title: "Statistical Inference Project - Part II"
subtitle: "Analysis  of the ToothGrowth data"
author: "Loïc BERTHOU"
date: "August 19, 2015"
output: pdf_document
---

```{r, echo=FALSE, results='hide'}
set.seed(20150819)
library(ggplot2)
```

# Overview

This document is a basic analysis of the ToothGroth dataset provided in the R datasets packages.

# Introduction

Since we don't have any knowledge about this dataset, we will first perform an exploratory analysis to get a better sense of the type of data we will be dealing with.
Then we will perform a comparison of the tooth groth by supp and dose as requested.

# Exploratory data analysis

The first thing to do with the dataset is to check if there is any sort of documentation attached to it. 

```{r}
help("ToothGrowth")
```

The _str_ function returns some useful information.

```{r}
str(ToothGrowth)
```

Finally, I will draw a first plot of the scattered points.

```{r}
g <- ggplot(data=ToothGrowth, aes(x=dose, y=len))
g <- g + geom_point(aes(color=supp), size=5, alpha=.5)
g
```

# Summary of the data

We have now a first idea of the data manipulated.
These numbers have been produced by a study on the effect of Vitamin C on guinea pigs and more precisely the length of their teeth.
The vitamin C has been administered either in the form of orange juice (represented "OJ"") or ascorbic acid (represented "VC"). Each form has been administered at 3 dose levels to 10 animals (60 different animals in total).

# Tooth growth comparison by supp and dose

TODO

# Conclusion

TODO
